---
title: 'Introduction to APIs'
author: "A. Paxton"
output:
  html_document:
    keep_md: yes
    number_sections: yes
---

An API---or *application programming interface*---is a formal avenue created 
by an organization or website to programmatically access their data. We could
most directly contrast an API with *web scraping*, which we will cover in a 
future session.

To use an API, you must submit a *request* to the system. That request
includes a set of information that has been specified by the API developers.
Minimally, the request will describe the kind of data that you want to access
(in the manner outlined by the API). For some systems, this must also include an
*API key*, which is an authenticated permission identifier that allows you to
access data. 

After submitting your request, you will receive a *response* that contains the
data you've requested. Your next job will be to parse the response into usable
data.

In this exercise, we'll walk through the process of using an API directly from R.

***

# Preliminaries

First, we'll need to prepare for our exercises. We'll do this by loading the
packages we need, including---if needed---installing those packages. For the
sake of the exercise, we'll load the packages silently, meaning that we won't
clutter our R markdown output with a bunch of warnings and output messages
from the loading and installation process.

```{r clear-workspace}

# clear the workspace (useful if we're not knitting)
rm(list=ls())

```

```{r function-check-for-packages, include=FALSE}

# make sure we can load packages 
# (thanks to https://gist.github.com/smithdanielle/9913897)
load_or_install_packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, 
                     dependencies = TRUE,
                     repos="http://cloud.r-project.org/")
  sapply(pkg, require, character.only = TRUE)
}

```

```{r load-packages, message=FALSE, results="hide"}

# specify which packages we'll need
required_packages = c("tidyverse",
                      "httr",
                      "jsonlite")

# install them (if necessary) and load them
load_or_install_packages(required_packages)

```

***

# Finding APIs

Not every website, repository, or organization has an API available for users to
access data, but those that do have them tend to make them easily available. You
might start by checking the website for an API information link. This might be
at the bottom of the webpage or available in a "Developers" menu of the website.
If you can't find one, try using the "find" function in your browser for `API`,
or try searching `API` in the website's search function (if they have one).

If you still can't find it, try doing a general search on your favorite search
engine with the name of the website and API. (As you may already know, putting
both terms in separate double-quotes will ensure you only get webpage results
that include both terms.)

If you still can't find anything, it may be time to turn to web scraping---but,
again, we won't be covering that here. (Again, keep in mind matters of
ethicality and legality, including the *hiQ Labs v. LinkedIn* ruling in the US.)

***

