---
title: 'Introduction to APIs'
author: "A. Paxton"
output:
  html_document:
    keep_md: yes
    number_sections: yes
---

An API---or *application programming interface*---is a formal avenue created 
by an organization or website to programmatically access their data. We could
most directly contrast an API with *web scraping*, which we will cover in a 
future session.

To use an API, you must submit a *request* to the system. That request
includes a set of information that has been specified by the API developers.
Minimally, the request will describe the kind of data that you want to access
(in the manner outlined by the API). For some systems, this must also include an
*API key*, which is an authenticated permission identifier that allows you to
access data. 

After submitting your request, you will receive a *response* that contains the
data you've requested. Your next job will be to parse the response into usable
data.

In this exercise, we'll walk through the process of using an API directly from R.

***

# Preliminaries

```{r function-check-for-packages, include=FALSE}

# make sure we can load packages 
# (thanks to https://gist.github.com/smithdanielle/9913897)
load_or_install_packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, 
                     dependencies = TRUE,
                     repos="http://cloud.r-project.org/")
  sapply(pkg, require, character.only = TRUE)
}

```

```{r preliminaries}

# specify which packages we'll need
required_packages = c("tidyverse",
                      "httr",
                      "jsonlite")

# install them (if necessary) and load them
load_or_install_packages(required_packages)

```

